---
title: "Rotorua Light"
subtitle: "Parameterising light extinction (Kw) for a GLM-AED climate scenario run"
format:
  html:
    toc: true
    code-fold: show
    df-print: kable
execute:
  warning: false
---

## Motivation

Lake Rotorua's GLM-AED setup needs a light extinction (`Kw`) parameterisation.
The options considered were:

1. A single **static Kw** (one value for the whole run).
2. A **monthly Kw climatology** (12 calendar-month values, time-varying).
3. A **fully dated time-varying Kw** built from the observed record.
4. A **decomposed background Kw** (observed Kw minus the fraction attributable
   to phytoplankton/turbidity), for use when AED's own biology should generate
   the biotic part of light extinction dynamically.

This document works through the diagnostics behind that decision for the
Rotorua PAR/Kd record, and lands on a recommendation for the **climate
scenario** runs specifically (as opposed to a historical hindcast).

The helper functions used throughout live in
[`R/recommend_light_parameterisation.R`](R/recommend_light_parameterisation.R).

```{r}
#| label: setup
#| echo: false
#| include: false
library(ggplot2)
library(patchwork)
library(AEME)
source("../R/recommend_light_parameterisation.R")
source("R/setup_aeme.R")
path <- "light_test"
aeme <- setup_aeme(path = path)
```

## Static vs. time-varying Kw: the trade-offs

**Time-varying Kw** (monthly or dated) can capture real seasonal drivers of
light attenuation — spring runoff/turbidity pulses, algal self-shading,
CDOM inputs, ice-off resuspension — that a single Kw cannot. Where good
profile data exist, throwing that signal away by collapsing to one number
misrepresents both the clear and the turbid periods.

The trade-offs:

- More data wrangling and another forcing file to QC.
- Monthly resolution still misses short-lived bloom/storm spikes.
- If Kw has been used as a tuning knob to compensate for other model
  deficiencies (mixing scheme, inflow temperatures), imposing an
  independently-derived time-varying Kw removes that degree of freedom and
  can expose calibration problems elsewhere.
- A poorly-constrained time-varying curve (built on noisy or sparse data) is
  worse than a well-chosen static value.

**Static Kw** is defensible when extinction is dominated by relatively
stable factors, when supporting data are too sparse/noisy to justify more
resolution, or — as turns out to matter most here — **when there is no
calendar correspondence between the simulated period and the observed
record**, which is exactly the case for a forward climate scenario.

## GLM-AED's own light feedback

When AED is coupled to GLM with `bioshade_feedback = .true.`, light
extinction is **not** just the prescribed `Kw` — GLM adds AED's own
simulated biotic/particulate extinction on top, additively
(`glm_aed.F90`, `update_light`):

```fortran
par(i) = par(i+1) * EXP( -(Kw + localext_up) * dz(i+1) )
IF (bioshade_feedback) extc(i) = Kw + localext
```

`localext` is the sum of whatever AED modules contribute a light-extinction
diagnostic — in this setup that includes:

- **`aed_phytoplankton`** — self-shading from simulated chlorophyll.
- **`aed_noncohesive`** — extinction from simulated suspended sediment,
  `Ke_ss(i) * SS_conc(i)` summed across size classes. The active
  configuration:

```fortran
&aed_noncohesive
   num_ss = 2
   ss_initial = 3, 1
   Ke_ss = 0.06, 0.06
   settling = 1
   w_ss = 0.5, 0.03
   rho_ss = 1800, 1200
   d_ss = 2e-05, 4e-06
   resuspension = 1
   epsilon = 0.02
   tau_0 = 0.04, 0.03
   tau_r = 1
   fs = 1, 1
   Fsed = 0, 0
   decay = 0, 0
   simSedimentMass = .false.
   sed_porosity = 0.3
   diag_level = 10
/
```

**Practical consequence:** the prescribed `Kw`/`Kw_file` should represent
only the *background* (non-biotic, non-suspended-sediment) extinction —
water + CDOM + anything not reproduced by AED's own state. If a Kw derived
from field-measured Kd (which already contains the biotic + SS signal at the
time of sampling) is combined with `bioshade_feedback` and an active
`aed_noncohesive`, the biotic/SS contribution is **double-counted**,
over-attenuating light and biasing production, DO, and thermal structure.

## Configuring GLM to use a time-varying Kw

For reference, if a time-varying background is used, the `&light` namelist
and `Kw_file` CSV are:

```fortran
&light
   light_mode       = 0        ! single-Kw Beer-Lambert; required for Kw_file
   Kw               = 0.37     ! fallback/initial value
   Kw_file          = 'kw.csv' ! time-varying background Kw forcing
   n_bands          = 2
   light_extc       = 1.0, 0.5
   energy_frac      = 0.5, 0.5
   Benthic_Imin     = 10
   albedo_mode      = 1
   albedo_mean      = 0.08
   albedo_amplitude = 0.02
/
```

```csv
time,Kd
2015-01-01,0.62
2015-02-01,0.58
2015-03-01,0.51
```

- The `time` column must use the same `time_fmt`/`timezone` as the
  `&meteorology` block.
- GLM performs a **daily lookup** (holds the most recent row's value), not
  sub-daily interpolation — the file needs one row per day you want a
  distinct value for.
- `Kw_file` only drives the single scalar `ExtcCoefSW`, so it requires
  `light_mode = 0`; it has no effect on the per-band `light_extc` array used
  by `light_mode = 1`.

## Diagnosing the observed Kd record

`kd_fits` holds Kd values fitted from PAR profiles (`log(PAR) ~ depth`,
filtered to r² > 0.8), spanning 2003–2025.

```{r}
#| label: load-light-data
#| include: false
targets::tar_config_set(store = here::here("_targets"))
targets::tar_load(light_data)
```

`light_data` provides the same `Date`/`Kd` fields used throughout this
analysis (fitted from PAR profiles via `log(PAR) ~ depth`, filtered to
r² > 0.8), spanning 2003–2025.

```{r}
#| label: light-data-summary
summary(light_data)
```

When Kd and secchi depth are plotted over time, the **interannual variability dominates over any
seasonal signal**:

```{r}
#| label: kd-fits-plot
#| fig-width: 9
#| fig-height: 4.5

p1 <- ggplot(light_data, aes(x = Date, y = Kd)) +
  geom_line() +
  geom_point(aes(colour = par_n)) +
  labs(x = "Date", y = "Kd (m⁻¹)", title = "Observed Kd over time") +
  scale_colour_viridis_c(name = "n PAR meas", option = "plasma") +
  theme_minimal()

p2 <- ggplot(light_data, aes(x = Date, y = secchi)) +
  geom_line() +
  geom_point() +
  labs(x = "Date", y = "Secchi depth (m)", title = "Observed Secchi depth over time") +
  theme_minimal()

g <- wrap_plots(p1, p2, ncol = 1)
g
```

There is not a consistent relationship between the CTD PAR profiles and the 
secchi depth:

```{r}
#| label: kd-v-secchi-plot
#| fig-width: 9
#| fig-height: 4.5

light_data <- light_data |> 
  dplyr::mutate(month = lubridate::month(Date, label = TRUE, abbr = TRUE))

ggplot(light_data, aes(x = secchi, y = Kd)) +
  geom_point(aes(colour = month)) +
  geom_smooth(method = "lm", se = TRUE, colour = "black") +
  labs(x = "Secchi depth (m)", y = "Kd (m⁻¹)", title = "Kd vs Secchi depth") +
  scale_colour_viridis_d(name = "Date", option = "plasma") +
  theme_bw()
```




```{r}
#| label: kd-fits-recommendation
kd_recommendation <- recommend_light_parameterisation(
  date = light_data$Date,
  kw   = light_data$Kd
)
print(kd_recommendation)
```

```{r}
#| label: kd-fits-plot
#| fig-width: 9
#| fig-height: 4.5
plot(kd_recommendation)
```

### Key finding: interannual variability dominates over seasonality

- Kd mean 0.71 m⁻¹, CV 0.52, range 0.25–1.92 m⁻¹.
- **Year explains ~47% of total variance** — more than the seasonal cycle
  does. Annual means step from ~0.35–0.43 m⁻¹ (2003–2008) to a noisy
  0.32–1.4 m⁻¹ regime from 2009 onward.
- After removing each year's own mean, the detrended seasonal ratio is only
  ~1.6–1.85x, and mostly driven by a handful of bloom/turbidity events rather
  than a robust recurring monthly pattern.
- A pooled multi-year **monthly climatology would blend a real interannual
  regime shift with a weak seasonal signal**, misrepresenting most individual
  years — hence the function's recommendation of a dated time-varying series
  over a monthly climatology *for a hindcast*.

## Decomposing observed Kd into background vs. biotic components

Available overlapping chemistry: `CHLA (mg/m^3)` and `Turbidity (NTU)` are
the two usable predictors of the biotic/particulate extinction component
(Secchi is itself just another optical measurement of the same quantity, so
it is not used as a predictor).

```{r}
#| label: decompose-kw-example
#| eval: true
# `chem` should be one row per Kd date (aggregate multiple depths/stations to
# a single near-surface value per date *before* this step — the first attempt
# at this joined on Date without deduplicating, inflating n from 147 to 841
# rows and producing an unreliable, pseudo-replicated fit).
kw_decomp <- decompose_kw_biotic(
  date      = light_data$Date,
  kw        = light_data$Kd,
  chla      = light_data$chla,
  turbidity = light_data$turbidity
)

```

```{r}
#| label: decompose-kw-summary
print(kw_decomp)
```


Two fits were run against the real data as the join was corrected:

```{r}
#| label: decompose-kw-plot
#| fig-width: 9
#| fig-height: 4.5
plot(kw_decomp)
```


```{r}
#| label: decompose-kw-summary
print(kw_decomp)
```


### Interpretation

- Even after fixing the join (pseudo-replication), **R² = 0.067** — Chla and
  turbidity together explain under 7% of Kd variance. Chla alone is not
  significant (p = 0.167); only turbidity clears p < 0.05, and its large
  coefficient combined with an 18% floor-clipping rate suggests the fit is
  being pulled around by a handful of high-turbidity events rather than
  reflecting a stable relationship.
- `aed_noncohesive` is active in this AED configuration, computing its own
  suspended-sediment-driven extinction (`Ke_ss(i) * SS_conc(i)`) dynamically
  — so, unlike a setup with phytoplankton feedback alone, **both** the Chla-
  and turbidity-attributable terms are legitimately things AED will
  reconstruct, and both are fair to remove from the prescribed background.
  (Caveat: `aed_noncohesive`'s simulated SS is a mass concentration state,
  not the same quantity as measured NTU turbidity — the *decision* to
  exclude it is justified, but there's no guarantee the simulated SS
  dynamics reproduce the same magnitude/timing the regression attributed to
  turbidity.)
- The **intercept is the one robust number here**: 0.368 m⁻¹ at p = 0.0013,
  much better constrained than either slope, since intercepts are anchored
  by the bulk of the data rather than a weak covariate relationship.
- The Secchi–Kd relationship was also noted as weak, reinforcing that this
  system's optical signal is noisy relative to the predictors available.

## Recommendation for the climate scenario runs

For a **climate scenario** parameterisation (not a historical hindcast):

1. **Do not use observed Kd.** `bioshade_feedback` and `aed_noncohesive` are
   both active, so a Kw already containing today's biotic/SS signal would be
   double-counted against AED's own dynamic contribution.
2. **Do not use a dated time-varying background.** There is no calendar
   correspondence between a historical sampling date and a future scenario
   year — the "use the actual dated observations" option only applies to
   hindcast-style runs over the sampled period.
3. **Do not build a monthly background climatology.** The Chla/turbidity
   regression (R² = 0.067) and the Secchi–Kd relationship are both weak;
   imposing monthly structure on this background signal risks encoding noise
   as if it were a genuine recurring seasonal driver.
4. **Use a static background Kw**, set from the regression intercept
   (~0.37 m⁻¹) — the most robust single number the data support. Let
   `bioshade_feedback` (phytoplankton) and `aed_noncohesive` (suspended
   sediment) supply the dynamic, climate-responsive part of the light
   environment, rather than freezing it to the historical record. This is
   also the conceptually correct setup for a scenario run: it lets the
   lake's optical response evolve mechanistically under whatever the
   scenario does to nutrients, temperature, and hydrology, instead of
   repeating today's bloom/turbidity patterns indefinitely.

```fortran
&light
   light_mode       = 0
   Kw               = 0.37   ! static background Kw (regression intercept)
   n_bands          = 2
   light_extc       = 1.0, 0.5
   energy_frac      = 0.5, 0.5
   Benthic_Imin     = 10
   albedo_mode      = 1
   albedo_mean      = 0.08
   albedo_amplitude = 0.02
/
```

## Sensitivity check

Before committing to this, a direct sensitivity comparison is worth running:
static background Kw (0.37 m⁻¹) with `bioshade_feedback` on, vs. the old
static *observed* mean Kw (0.71 m⁻¹) with `bioshade_feedback` off. If key
scenario outputs (surface temperature, stratification onset/breakdown,
chlorophyll, hypolimnetic DO) are not materially different between the two,
the choice matters less than the analysis above might suggest; if they
diverge substantially, it's worth investing more in constraining the
background estimate (e.g. more Chla/TSS sampling, a better SS-turbidity
calibration for `aed_noncohesive`) before running the full scenario suite.

The chunks below are scaffolding only (`eval: false`) — wire in your own nml
paths, `glm.exe` invocation, and output-reading (e.g. `glmtools`/`ncdf4`) to
run them.

```{r}
#| label: sensitivity-setup
#| eval: true

lake_dir <- get_lake_dir(aeme)
model <- "glm_aed"
model_path <- file.path(lake_dir, model) # get_model_path(aeme)
tgt_vars <- c("HYD_temp", "PHY_tchla", "LKE_photic", "HYD_thmcln")
scenarios <- list(
  `Static - feedback` = list(
    Kw = 0.37, # static background Kw (regression intercept)
    bioshade_feedback = TRUE
  ),
  `Static - no feedback` = list(
    light_mode = 0,
    Kw = 0.71, # static observed mean Kw
    bioshade_feedback = FALSE
  )
)

```

```{r}
#| label: sensitivity-run
#| eval: true

out <- lapply(names(scenarios), function(scn) {
  run_model_test(model = model, path = model_path, param_overrides = scenarios[[scn]], 
                 tgt_vars = tgt_vars)
})
names(out) <- names(scenarios)

```

```{r}
#| label: sensitivity-compare-temp
#| eval: true

plot_list <- lapply(tgt_vars, \(v) {
  var_lims <- range(unlist(lapply(out, function(lst) {
    range(lst[[v]], na.rm = TRUE)
  })), na.rm = TRUE)
  if (v %in% c("LKE_photic", "HYD_thmcln")) {
    df <- lapply(out, function(lst) {
      data.frame(Date = lst$Date, value = lst[[v]])
    }) |> 
      dplyr::bind_rows(.id = "scenario")
    p <- ggplot() +
      geom_line(data = df, aes(x = Date, y = value, colour = scenario)) +
      labs(x = "Date", y = v) +
      scale_colour_viridis_d(name = "Scenario") +
      theme_minimal() +
      scale_y_reverse(limits = c(0, max(var_lims)))
    return(p)
  }
  plist <- lapply(names(out), function(scn) {
    lst <- out[[scn]]
    plot_model_output(lst, var_sim = v, var_lims = var_lims) +
      ggtitle(label  = paste("Scenario:", scn))
  })
  wrap_plots(plist, ncol = 1) +
    plot_annotation(title = paste("Comparison of", v, "between scenarios"))
})
names(plot_list) <- tgt_vars


```

```{r}
#| label: sensitivity-plot-temp
#| eval: true
#| fig-width: 9
#| fig-height: 4.5
plot_list$HYD_temp
```

```{r}
#| label: sensitivity-plot-photic-depth
#| eval: true
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Photic depth (m) comparison between scenarios"
plot_list$LKE_photic
```

## Summary

| Question | Answer |
|---|---|
| Static vs. time-varying Kw (general)? | Time-varying is better *if* the seasonal signal is real and well-sampled; here, interannual variability (47% of variance) dominates over seasonality, so a pooled monthly climatology would misrepresent most years. |
| Does AED feedback interact with Kw? | Yes — `bioshade_feedback` (phytoplankton) and `aed_noncohesive` (suspended sediment) both add their own simulated extinction on top of the prescribed `Kw`, additively. |
| Should prescribed Kw include the biotic signal? | No, not when both feedbacks are active — it double-counts. |
| Can the biotic component be estimated from data? | Weakly. `Kd ~ Chla + Turbidity` gives R² = 0.067; only the intercept (background Kw ≈ 0.37 m⁻¹) is well-constrained. |
| Is calibrating AED's biotic light coefficients a good idea? | Not as a first move — risks equifinality/circularity with growth and mortality parameters; better to fix it from the empirical estimate and reserve calibration for parameters not otherwise measurable. |
| What should the climate scenario runs use? | A **static background Kw** (~0.37 m⁻¹), with `bioshade_feedback` and `aed_noncohesive` supplying the dynamic, climate-responsive part of light extinction. |
