---
title: "Meteorological Data Sources & Forcing Comparison"
format: html
editor: visual
warning: false
message: false
---

```{r}
#| label: setup
#| include: false
#| echo: false

library(targets)
library(dplyr)
library(ggplot2)
```

This page compares the four meteorological data sources available for Lake Rotorua, describes what each one brings to the model forcing built in [`climate-scenario-workflow.qmd`](climate-scenario-workflow.qmd), and documents a couple of data-quality issues this comparison caught along the way. See [`data.qmd`](data.qmd) for full source/access details on every dataset in the pipeline, including these four.

## The four sources

| Source | Target(s) | Record | What it measures | Where it's strong | Where it's weak |
|----|----|----|----|----|----|
| **Lake buoy** | `rotorua_buoy_met_aeme_hr` | 2022–present (\~4.5 yr) | On-lake met station: temp, RH, wind, radiation, rain | The only source actually measured *over the lake surface* — the ground truth for lake-effect corrections | Short record; rain gauge is noisy (weak correlation with land stations, see below) |
| **Airport (NIWA #1770)** | `niwa_met_hourly_aeme` | 1981–2025 (44 yr) | Full met suite at a fixed land station | Long, high-quality, consistent instrumentation | Land-station bias vs. the lake (colder, drier air; needs the buoy transfer function to be lake-representative) |
| **ERA5-Land (reanalysis)** | `era5_hourly_met` | 2011–present | Gridded reanalysis, full met suite | Continuous, gap-free, extends into the future for scenario work | Coarse grid cell (not lake-specific); its own biases need correcting against real observations |
| **PC10 / BOPRC stations** | `pc10_climate_met` | Temp/RH: 2006–present (Edmund Rd); Rain: **1901–present** (Whakarewarewa) | Temp, RH, rain only — no radiation or wind | Rain record is by far the longest of any source here; independent of NIWA's instrumentation | No radiation/wind; weaker correlation with the buoy than the airport (see comparison below) |

The airport and PC10 stations were chosen from BOPRC/NIWA's larger station networks by longest, most complete overlapping record rather than by raw distance to the lake — see [the PC10 extraction notes](targets.qmd) for the full station comparison (`pc10_climate_station_summary()`); for Lake Rotorua, every realistic candidate station already sits close to the lake, so record length matters more than the last kilometre of proximity.

## Comparing the raw sources against the buoy

`met_source_comparison` aggregates all four sources to daily values and scores the airport, ERA5 and PC10 against the buoy — the only source with a direct claim to being "true" lake conditions — over whatever window each pair overlaps (2022 onward, bounded by the buoy).

```{r}
#| label: comparison-stats
#| echo: false

cmp <- tar_read(met_source_comparison)
cmp$stats |>
  knitr::kable(digits = 3,
              caption = "Daily bias/MAE/RMSE/r of each source vs. the buoy, computed only over days both series have valid data.")
```

```{r}
#| label: comparison-plot
#| echo: false
#| fig-height: 8
#| fig-cap: "Daily air temperature, relative humidity and rainfall from all four sources, over the buoy's record."

tar_read(met_source_comparison_plot)
```

**What this shows:**

-   **Air temperature** — every land/reanalysis source reads \~1–1.5 °C colder than the buoy, which is the expected lake-thermal-inertia effect (water buffers diurnal and seasonal swings that land stations don't), not a data error. ERA5 tracks the buoy's day-to-day shape best (r ≈ 0.99) despite the largest mean offset; PC10 has the weakest correlation (r ≈ 0.85) of the three.
-   **Relative humidity** — all sources read more humid than the buoy on average. Correlation is much weaker across the board than for temperature, and **PC10's RH correlation with the buoy is very weak (r ≈ 0.28)** — treat PC10 RH as a rough guide, not a precise match to lake conditions.
-   **Rainfall** — airport and PC10 both track the buoy's rain total closely on average (bias ≈ 0). PC10's *day-to-day* correlation with the buoy is weak (r ≈ 0.12), consistent with either real small-scale spatial variability in rainfall (rain is far patchier than temperature) or gauge-timing noise, rather than a systematic error — its long-run total still agrees.

## A real bug this comparison caught

Running this comparison surfaced a unit mismatch that a single-source sanity check would have missed: **ERA5-Land's raw `MET_pprain` column is stored in mm/hour, while every other source in this pipeline (buoy, airport, PC10) is in m/hour/m-day**, the AEME convention. Uncorrected, ERA5's rainfall was reading roughly 1000x too high in the shared units (mean 0.18 vs. the buoy's 0.0008 m/hr) — and because `era5_bias_correction` originally excluded `MET_pprain` from correction (rainfall was assumed to be a poor target due to buoy-gauge reliability, not a units problem), that inflated value would have flowed straight through `era5_corrected_hourly` into every scenario baseline.

Fixed in two places:

1.  **[`read_era5_hourly_met()`](https://github.com/tadhg-moore/rotorua-pc10/blob/main/R/scenario_climate_workflow.R)** now converts `MET_pprain` mm/hour → m/hour at read time, so the fix applies everywhere the ERA5 series is used, not just here.
2.  **`MET_pprain` is now included** in both `era5_bias_correction` and `era5_bias_correction_extended`'s `vars`, using the same `method = "scale"` day-of-year correction as the other variables — see the comparison above for why this is reasonable (no evidence of a buoy under-catch, just noise), while still treating the buoy's rain series with more caution than its temperature/RH/wind.

```{r}
#| label: era5-pprain-skill
#| echo: false

bc <- tar_read(era5_bias_correction)
bc$skill |>
  filter(variable == "MET_pprain", stage %in% c("cv_raw", "cv_corrected")) |>
  mutate(kind = ifelse(grepl("raw", stage), "raw ERA5", "bias-corrected")) |>
  select(variable, kind, n, bias, mae, rmse, r) |>
  knitr::kable(digits = 4,
              caption = "Leave-one-year-out cross-validated skill for MET_pprain specifically, before vs. after the scale correction (post unit-fix).")
```

## How this improves lake model forcing

1.  **Caught a forcing bug before it reached the model.** The ERA5 rain unit error above would have meant every ERA5-driven scenario — including every future GCM projection, since deltas are applied to the ERA5 baseline — ran with rainfall roughly 3 orders of magnitude too high. A single-source pipeline has no way to notice this; comparing independent sources does.
2.  **Validates the existing buoy → airport extension.** `climate-scenario-workflow.qmd` already extends the short (\~4.5 yr) buoy record using the 44-year airport record (`niwa_to_buoy_bias_correction` → `niwa_met_hourly_lake_equiv`) so `era5_bias_correction_extended` has a longer, still lake-representative reference. PC10's independent record over the same window is a natural cross-check on that extension, since it comes from different instrumentation and a different network (BOPRC, not NIWA).
3.  **PC10's 1901–present rainfall record is the longest available context for Rotorua rainfall**, over 40 years longer than the airport's. That's directly useful for `summarise_climate_extremes.R` / [`climate-extreme-events.qmd`](climate-extreme-events.qmd)'s historical-extremes framing, and as a long-run independent check on whatever ERA5 rainfall climatology ends up driving the historical baseline and scenario deltas.
4.  **Flags where to be careful.** PC10's weak correlation with the buoy for RH and rainfall (r ≈ 0.28 and 0.12) means it shouldn't be used as a `fit_met_bias_correction()` reference for those variables the way the airport is — its value here is historical context and cross-validation, not as a second forcing bridge.

## Caveats carried over from `climate-scenario-workflow.qmd`

-   **Station pressure (`MET_prsttn`)** has a suspected unit issue in `buoy_to_aeme_met()` (raw bias vs. ERA5 \~96 kPa, an order of magnitude too large to be a real elevation effect) — not yet resolved, unrelated to the rainfall fix above.
-   PC10 has **no radiation or wind data**, so `pc10_climate_met` is a partial forcing (temp/RH/rain only) — useful for comparison and historical context, not as a standalone forcing source.
