---
title: "Lake Rotorua V4 - GLM-AED output plots"
format: html
eval: true
execute:
  cache: true
knitr:
  opts_chunk:
    cache.extra: !expr tools::md5sum(c("LakeRotorua/glm4_original.nml", "LakeRotorua/aed/aed_original.nml"))
---

Rotorua V4 outputs (GLM-AED+ 4.0.0, `glm4.nml`, 1-year run 2008-07-01 to 2009-07-01),
plotted with the [AEME](https://github.com/limnotrack/AEME) package rather than a
hand-rolled `netCDF4`/`matplotlib` stack — this is the R/Quarto counterpart of the old
`plot_rotorua.ipynb`.

`AEME::get_var()`/`plot_output()` cover every plain water-column and sediment-zone
variable (they read `output/output.nc` and `output/lake.csv` for you, translate between
AED and AEME variable names, and turn a depth-time array into a heatmap in one call). The
one thing they don't model is the dynamic sediment-heat profile (`sed_heat_model = 2`,
a `(time, nzones, sed_layers)` array) — that section still opens the NetCDF file directly
with `ncdf4`.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(AEME)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(ncdf4)

lake_dir  <- "LakeRotorua"
output_nc <- file.path(lake_dir, "output", "output.nc")
plot_dir  <- file.path(lake_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE)
dir.create(file.path(plot_dir, "profile"), showWarnings = FALSE)
dir.create(file.path(plot_dir, "zone"), showWarnings = FALSE)

# glm_exec <- normalizePath(list.files(pattern = "*.exe", full.names = TRUE, recursive = TRUE))
# options("AEME.glm_exec" = glm_exec)

# file.copy(file.path(lake_dir,"aed", "aed_original.nml"),
#           file.path(lake_dir, "aed", "aed.nml"), overwrite = TRUE)
aeme <- glm_config_to_aeme(nml_file = file.path(lake_dir, "glm4_original.nml"))

tgt_vars <- c("HYD_temp", "CHM_salt", "HYD_dens", "RAD_extc", "OXY_oxy", "CAR_dic", "CAR_pH",
              "PHS_frp", "ALU_ala", "ALU_alp", "ALU_uptake_w", "ALU_fpH",
              "ALU_ben_ala", "ALU_ben_alp", "ALU_uptake_s")

model_controls <- get_model_controls(use_bgc = TRUE)
model_controls <- set_vars_sim(model_controls, tgt_vars)

aeme <- aeme |> 
  build_aeme(use_bgc = TRUE, use_aeme = F)
aeme_path <- get_lake_dir(aeme)

aed_old_file <- file.path(lake_dir, "aed", "aed_original.nml")
aed_file <- file.path(aeme_path, "glm_aed", "aed", "aed.nml")

aed_old <- read_nml(aed_old_file)
aed <- read_nml(aed_file)
aed$aed_models$models <- c(aed$aed_models$models, "aed_alum")
aed$aed_alum <- aed_old$aed_alum
write_nml(aed, aed_file)

aeme <- aeme |> 
  run_aeme()

outfile <- get_model_outfile(aeme)$glm_aed["output"]
out <- read_glm_output(file = outfile, vars_sim = tgt_vars, load_all = T)

theme_set(theme_bw(base_size = 10))
```

## 0. Helpers

`profile_vars`/`zone_vars` are read straight off the NetCDF file's dimensions (AEME has
no catalogue of "every variable in this output" — it only knows the variables it has a
name mapping for), so this is the one place raw `ncdf4` is used outside the sediment-heat
section. `get_var()` happily accepts either the AED name (`"OXY_oxy"`) or AEME's own name
(`"CHM_oxy"`) for everything found this way.

```{r}
#| label: helpers
#| message: false

nc <- nc_open(outfile)
var_dims <- sapply(names(nc$var), function(v) sapply(nc$var[[v]]$dim, `[[`, "name"))

profile_vars <- names(var_dims)[sapply(var_dims, function(d) "z" %in% d)]
zone_vars    <- names(var_dims)[sapply(var_dims, function(d) {
  "nzones" %in% d && !"z" %in% d && !"sed_layers" %in% d
})]

long_name <- function(var) {
  v <- nc$var[[var]]
  if (is.null(v)) return(var)
  sprintf("%s (%s)", if (nzchar(v$longname)) v$longname else var, v$units)
}

cat(sprintf("%d profile variables, %d zone variables\n", length(profile_vars), length(zone_vars)))

# A handful of AED variable names aren't recognised by AEME's fuzzy matcher even
# though they're just renamed internally (e.g. dissolved oxygen is `CHM_oxy`, not
# `OXY_oxy`) - resolve those explicitly rather than letting get_var() error.
var_alias <- c(OXY_oxy = "CHM_oxy")
resolve_var <- function(var) unname(ifelse(var %in% names(var_alias), var_alias[var], var))

#' GLM writes each layer against a moving Lagrangian grid, so `get_var()`'s depths
#' don't line up day to day - interpolate onto a common depth-below-surface grid
#' before tiling, or the heatmap comes out as sparse, misaligned slivers.
grid_profile <- function(df, nz = 150) {
  depths <- seq(0, max(df$depth, na.rm = TRUE), length.out = nz)
  df |>
    group_by(Date) |>
    group_modify(~ data.frame(depth = depths,
                               value = approx(.x$depth, .x$value, xout = depths, rule = 2)$y)) |>
    ungroup()
}

#' Depth-time heatmap for a (lon, lat, z, time) variable, via AEME::get_var()
plot_profile <- function(var, cmap = "viridis") {
  df <- get_var(aeme, model = "glm_aed", var_sim = resolve_var(var), remove_spin_up = FALSE) |>
    grid_profile()
  ggplot(df, aes(x = Date, y = depth, fill = value)) +
    geom_raster() +
    scale_y_reverse() +
    scale_fill_viridis_c(option = cmap, na.value = NA) +
    labs(x = NULL, y = "depth (m)", fill = NULL, title = var,
         subtitle = long_name(var)) +
    theme(plot.title = element_text(size = 10), plot.subtitle = element_text(size = 8))
}

#' One line per sediment zone for a (lon, lat, nzones, time) variable
plot_zone <- function(var) {
  df <- get_var(aeme, model = "glm_aed", var_sim = resolve_var(var), remove_spin_up = FALSE)
  ggplot(df, aes(x = Date, y = value, colour = factor(nzones))) +
    geom_line(linewidth = 0.4) +
    labs(x = NULL, y = NULL, colour = "zone", title = var, subtitle = long_name(var)) +
    theme(plot.title = element_text(size = 9), plot.subtitle = element_text(size = 7),
          legend.position = "bottom")
}
```

## 1. Lake-level summary

`plot_wlev()` and `plot_wbal()` are AEME's built-ins for the two standard GLM diagnostics;
the temperature and heat-balance panels are assembled from `get_var()` since AEME has no
single "lake summary" plot for them.

```{r}
#| label: lake-summary
#| fig-width: 12
#| fig-height: 8

temp_panel <- bind_rows(
  get_var(aeme, model = "glm_aed", var_sim = "surface_temp") |> mutate(series = "Surface Temp"),
  get_var(aeme, model = "glm_aed", var_sim = "max_temp")     |> mutate(series = "Max Temp"),
  get_var(aeme, model = "glm_aed", var_sim = "min_temp")     |> mutate(series = "Min Temp")
) |>
  ggplot(aes(x = Date, y = value, colour = series)) +
  geom_line(linewidth = 0.5) +
  scale_colour_manual(values = c("Surface Temp" = "firebrick", "Max Temp" = "darkorange",
                                  "Min Temp" = "steelblue")) +
  labs(title = "Surface / max / min temperature (degC)", x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "bottom")

heat_panel <- bind_rows(
  get_var(aeme, model = "glm_aed", var_sim = "LKE_Qsw") |> mutate(series = "Qsw"),
  get_var(aeme, model = "glm_aed", var_sim = "LKE_Qe")  |> mutate(series = "Qe"),
  get_var(aeme, model = "glm_aed", var_sim = "LKE_Qh")  |> mutate(series = "Qh"),
  get_var(aeme, model = "glm_aed", var_sim = "LKE_Qlw") |> mutate(series = "Qlw")
) |>
  ggplot(aes(x = Date, y = value, colour = series)) +
  geom_line(linewidth = 0.4) +
  labs(title = "Heat balance components", x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "bottom")

lake_summary <- (plot_wlev(aeme, model = "glm_aed") | plot_wbal(aeme, model = "glm_aed")) /
  (temp_panel | heat_panel)
lake_summary
ggsave(file.path(plot_dir, "lake_summary.png"), lake_summary, width = 12, height = 8, dpi = 120)
```

## 2. Physical profiles

```{r}
#| label: profile-headline
#| fig-width: 10
#| fig-height: 13

headline <- intersect(c("temp", "salt", "dens", "extc"), profile_vars)
profile_headline <- wrap_plots(lapply(headline, plot_profile), ncol = 1)
profile_headline[[1]] <- profile_headline[[1]] + scale_fill_viridis_c(option = "magma")
profile_headline
ggsave(file.path(plot_dir, "profile_headline.png"), profile_headline, width = 10, height = 13, dpi = 120)
```

## 3. AED biogeochemistry

The state-variable list is whatever `aed/aed.nml` currently simulates (`aed_sedflux`,
`aed_oxygen`, `aed_carbon`, `aed_phosphorus`, `aed_alum`) — a smaller set than the full
biogeochemistry the original notebook was written against, so the sample below is drawn
from what's actually in `profile_vars`.

```{r}
#| label: profile-aed-sample
#| fig-width: 10
#| fig-height: 12

aed_sample <- intersect(c("OXY_oxy", "CAR_dic", "CAR_pH", "PHS_frp"), profile_vars)
profile_aed_sample <- wrap_plots(lapply(aed_sample, plot_profile), ncol = 1)
profile_aed_sample
ggsave(file.path(plot_dir, "profile_aed_sample.png"), profile_aed_sample, width = 10, height = 12, dpi = 120)
```

## 4. Dynamic sediment heat model

`sed_heat_model = 2` gives every sediment zone a vertical soil-temperature profile,
integrated by `zZSoilTemp` with the zone-average water temperature as its surface boundary
condition. This is a `(time, nzones, sed_layers)` array with no AEME variable mapping, so
it's read directly from the NetCDF file.

What to look for: the seasonal signal should **damp and lag** with depth, and the deepest
node should be flat.

```{r}
#| label: sediment-temperature
#| fig-width: 11
#| fig-height: 10

sed_temp  <- ncvar_get(nc, "sed_temp")        # (sed_layers, nzones, time)
sed_depth <- ncvar_get(nc, "sed_layer_depth") # (sed_layers)
zone_ztemp <- ncvar_get(nc, "zone_ztemp")     # (nzones, time)
sed_hf     <- ncvar_get(nc, "sed_heatflux")   # (nzones, time)
time_vals  <- as.Date(ncvar_get(nc, "time") / 24, origin = strsplit(ncatt_get(nc, "time", "units")$value, "since ")[[1]][2])
n_zone     <- dim(sed_temp)[2]

cat("sediment nodes (m below SWI):", round(sed_depth, 3), "\n")

sed_df <- expand.grid(layer = seq_len(dim(sed_temp)[1]), zone = seq_len(n_zone), t = seq_along(time_vals)) |>
  mutate(Date = time_vals[t], depth = sed_depth[layer], value = sed_temp[cbind(layer, zone, t)])
water_df <- expand.grid(zone = seq_len(n_zone), t = seq_along(time_vals)) |>
  mutate(Date = time_vals[t], value = zone_ztemp[cbind(zone, t)])

sed_temp_plot <- ggplot(sed_df, aes(x = Date, y = value, colour = sprintf("%.2f m", depth))) +
  geom_line(linewidth = 0.4) +
  geom_line(data = water_df, aes(colour = "water"), linewidth = 0.9, colour = "black") +
  facet_wrap(~ sprintf("zone %d", zone - 1), ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "degC", colour = NULL, title = "Sediment temperature profile by zone") +
  theme(legend.position = "bottom")
sed_temp_plot
ggsave(file.path(plot_dir, "sediment_temperature.png"), sed_temp_plot, width = 11, height = 3.2 * n_zone, dpi = 120)
```

```{r}
#| label: sediment-heatflux
#| fig-width: 11
#| fig-height: 6

hf_df <- expand.grid(zone = seq_len(n_zone), t = seq_along(time_vals)) |>
  mutate(Date = time_vals[t], hf = sed_hf[cbind(zone, t)],
         diff = sed_temp[cbind(1, zone, t)] - zone_ztemp[cbind(zone, t)])

hf_plot <- ggplot(hf_df, aes(x = Date, y = hf, colour = factor(zone - 1))) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_line(linewidth = 0.4) +
  labs(title = "Sediment-water heat flux (W/m2; negative = water losing heat to the bed)",
       x = NULL, y = NULL, colour = "zone")

diff_plot <- ggplot(hf_df, aes(x = Date, y = diff, colour = factor(zone - 1))) +
  geom_line(linewidth = 0.4) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  labs(title = "Sediment-water temperature difference (degC)", x = NULL, y = NULL, colour = "zone")

sed_heatflux <- hf_plot / diff_plot
sed_heatflux
ggsave(file.path(plot_dir, "sediment_heatflux.png"), sed_heatflux, width = 11, height = 6, dpi = 120)
```

## 5. Alum (`aed_alum`)

`aed/aed.nml` currently sets `S_ala = 0` (no alum dose) with zero initial pools, so this
panel is mainly a check that the module is wired in — the min/max table below should stay
near zero and `ALU_fpH` near 1 (the pH modifier disabled) until `S_ala` is set to actually
dose the lake.

```{r}
#| label: alum-profiles
#| fig-width: 10
#| fig-height: 10

alu_prof  <- intersect(c("ALU_ala", "ALU_alp", "ALU_uptake_w", "ALU_fpH"), profile_vars)
alu_sheet <- intersect(c("ALU_ben_ala", "ALU_ben_alp", "ALU_uptake_s"), names(nc$var))

alum_profiles <- wrap_plots(lapply(alu_prof, plot_profile), ncol = 1)
alum_profiles
ggsave(file.path(plot_dir, "alum_profiles.png"), alum_profiles, width = 10, height = 10, dpi = 120)
```

```{r}
#| label: alum-benthic
#| fig-width: 11
#| fig-height: 4

alu_sheet_df <- lapply(alu_sheet, function(v) {
  get_var(aeme, model = "glm_aed", var_sim = v, remove_spin_up = FALSE) |>
    mutate(series = sprintf("%s  %s", v, long_name(v)))
}) |> bind_rows()

alum_benthic <- ggplot(alu_sheet_df, aes(x = Date, y = value, colour = series)) +
  geom_line(linewidth = 0.5) +
  labs(title = "Benthic alum pools", x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "bottom", legend.text = element_text(size = 7))
alum_benthic
ggsave(file.path(plot_dir, "alum_benthic.png"), alum_benthic, width = 11, height = 4, dpi = 120)

sapply(c(alu_prof, alu_sheet), function(v) {
  vals <- ncvar_get(nc, v)
  c(min = min(vals, na.rm = TRUE), max = max(vals, na.rm = TRUE))
}) |> t() |> knitr::kable(digits = 6)
```

## 6. Save every profile variable to PNG

One figure per variable under `plots/profile/`. Existing files are skipped, so re-running
is cheap; delete the directory to force a rebuild.

```{r}
#| label: dump-profile
#| message: false
#| warning: false

profile_dir <- file.path(plot_dir, "profile")
failed <- list()
for (var in profile_vars) {
  out_path <- file.path(profile_dir, paste0(var, ".png"))
  if (file.exists(out_path)) next
  p <- tryCatch(plot_profile(var), error = function(e) e)
  if (inherits(p, "error")) {
    failed[[var]] <- conditionMessage(p)
  } else {
    ggsave(out_path, p, width = 10, height = 4, dpi = 110)
  }
}
cat(sprintf("wrote %d profile PNGs to %s\n", length(profile_vars) - length(failed), profile_dir))
if (length(failed) > 0) print(failed)
```

## 7. Save every zone variable to PNG

Zone variables are `(nzones, time)` — sediment fluxes and the sediment heat flux.
`sed_temp` is handled separately in section 4 because it carries the extra `sed_layers`
dimension.

```{r}
#| label: dump-zone
#| message: false
#| warning: false

zone_dir <- file.path(plot_dir, "zone")
failed_z <- list()
for (var in zone_vars) {
  out_path <- file.path(zone_dir, paste0(var, ".png"))
  if (file.exists(out_path)) next
  p <- tryCatch(plot_zone(var), error = function(e) e)
  if (inherits(p, "error")) {
    failed_z[[var]] <- conditionMessage(p)
  } else {
    ggsave(out_path, p, width = 10, height = 4, dpi = 110)
  }
}
cat(sprintf("wrote %d zone PNGs to %s\n", length(zone_vars) - length(failed_z), zone_dir))
if (length(failed_z) > 0) print(failed_z)
```

## 8. Done

Everything is under [plots/](LakeRotorua/plots/):

- [plots/lake_summary.png](LakeRotorua/plots/lake_summary.png)
- [plots/profile_headline.png](LakeRotorua/plots/profile_headline.png), [plots/profile_aed_sample.png](LakeRotorua/plots/profile_aed_sample.png)
- [plots/sediment_temperature.png](LakeRotorua/plots/sediment_temperature.png), [plots/sediment_heatflux.png](LakeRotorua/plots/sediment_heatflux.png) — the dynamic sediment heat model
- [plots/alum_profiles.png](LakeRotorua/plots/alum_profiles.png), [plots/alum_benthic.png](LakeRotorua/plots/alum_benthic.png) — `aed_alum`
- [plots/profile/](LakeRotorua/plots/profile/) and [plots/zone/](LakeRotorua/plots/zone/) — one PNG per variable
