---
title: "Lake Rotorua — GLM Configuration"
subtitle: "General Lake Model (GLM 3.3.3) setup and simulation outputs"
author: "Tadhg Moore"
date: last-modified
date-format: "DD MMMM YYYY"
fig-width: 9
fig-height: 5
fig-dpi: 150
---

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(scales)
library(patchwork)
library(AEME)
library(knitr)
library(kableExtra)

source("../R/plot_aeme_loads.R")

glm_ver <- get_model_version("glm_aed")
glm_version <- regmatches(glm_ver[3], regexpr("\\d+\\.\\d+\\.\\d+", glm_ver[3]))
aeme <- readRDS("../LID11133_rotorua/aeme_download.rds")
cfg <- configuration(aeme)
glm3_nml <- cfg[["glm_aed"]][["hydrodynamic"]]
aed_nml <- cfg[["glm_aed"]][["bgc"]][["aed"]]

phyto_cols <- aed_nml$aed_phytoplankton$the_phytos + 1

inflows <- get_inflows(aeme)
inflow_humurana <- inflows[["Hamurana"]]

inp <- input(aeme)
met <- readr::read_csv("../LID11133_rotorua/glm_aed/bcs/meteo_glm.csv")
phyto_pars <- AEME::read_aed_param_csv("../LID11133_rotorua/glm_aed/aed/aed_phyto_pars.csv")
phyto_pars <- phyto_pars[, c(1, phyto_cols)]

# Model performance
aeme_time <- time(aeme)
period <- paste0(format(aeme_time$start, "%Y"), "–", format(aeme_time$stop, "%Y"))

perf <- assess_model(aeme, var_sim = c("HYD_temp","NIT_tn","PHS_tp","PHY_tchla","LKE_lvlwtr")) |> 
  dplyr::mutate(Period = period) |>
  dplyr::rename(Variable = name_text, RMSE = rmse, Bias = bias, R2 = r2) |> 
  dplyr::select(Variable, Period, RMSE, Bias, R2, n)

```

## Overview {#overview}

This page documents the **General Lake Model (GLM `r glm_version`)** configuration used for simulating the lake's thermal dynamics, water balance, and nutrient cycling. The model is coupled to the **Aquatic EcoDynamics (AED)** library for biogeochemical processes.

The [AEME](https://limnotrack.com/AEME/) package was used to retrieve and format lake characteristics, boundary conditions, and monitoring data for model configuration and validation. The `glm3.nml` and `aed.nml` files define the physical and biogeochemical model parameters, respectively, while CSV files provide time series of inflows, meteorological forcing, and phytoplankton parameters.

::: {.callout-note}
All simulation configuration files, scripts, and outputs documented here correspond to the
**baseline run** (`run_id: aeme_base`). 
:::

### Lake characteristics

```{r lake-summary-table}
#| echo: false
lake_char <- aeme |> 
  get_lake() |> 
  dplyr::mutate(
    area = area / 1e6, # convert from m2 to km2
  ) |> 
  dplyr::rename(
    Name = name, 
    ID = id,
    Lat = latitude,
    Lon = longitude,
    `Elevation (m)` = elevation,
    `Depth (m)` = depth,
    `Area (km2)` = area
  ) 

lake_char |>
  kbl(digits = c(0, 0, 4, 4, 1, 1), caption = "Summary of Lake Rotorua characteristics") |>
  kable_styling(bootstrap_options = c("striped","condensed","hover"),
                full_width = FALSE)


```

---

## Model schematic {#schematic}

The schematic below shows the overall GLM–AED model configuration for Lake Rotorua,
including the principal inflows, outflow, atmospheric forcing, and the biogeochemical
modules activated in this simulation.


```{r render-html, echo=FALSE}
plot_glm_config(aeme = aeme, return_widget = TRUE)
```

::: {.callout-tip collapse="true"}
## Schematic notes

- **Inflows**: Four principal inflows are configured (Hamurana, Ohau, Waiowhiro, Waiteti streams).
  Loads are derived from monitored flow and water quality data (BoPRC).
- **Outflow**: The Ohau Channel is the sole surface outflow; weir crest elevation set at 279.38 m a.s.l.
- **Meteorology**: Half-hourly data from the Rotorua Airport NIWA station (agent 89894), gap-filled
  using ERA5 reanalysis where needed.
- **WWTP**: Treated effluent from Rotorua WWTP is included as a point-source inflow.
- **AED modules active**: NIT (nitrogen), PHO (phosphorus), OGM (organic matter), PHY (phytoplankton).
:::

---

## Configuration files {#config-files}

:::{.panel-tabset}

### `glm3.nml`

::: {.callout-note appearance="minimal"}
Core GLM namelist — controls simulation period, physical parameters, mixing scheme, and I/O.
:::

```{=html}
<details open>
<summary style="cursor:pointer; font-weight:600; margin-bottom:0.5em;">
  glm3.nml — full file
</summary>
```

```{r}
#| echo: false
#| label: glm3-nml-preview

glm3_nml

```

```{=html}
</details>
```

### `aed.nml`

::: {.callout-note appearance="minimal"}
AED biogeochemical namelist — enables and configures nutrient, oxygen, organic matter, and phytoplankton modules.
:::

```{=html}
<details open>
<summary style="cursor:pointer; font-weight:600; margin-bottom:0.5em;">
  aed.nml — full file
</summary>
```

```{r}
#| echo: false
#| label: aed-nml-preview

aed_nml

```

```{=html}
</details>
```

### `inflow_Hamurana.csv`

::: {.callout-note appearance="minimal"}
Boundary condition file for Hamurana Stream — daily flow, temperature, and nutrient concentrations.
:::

```{=html}
<details>
<summary style="cursor:pointer; font-weight:600; margin-bottom:0.5em;">
  inflow_hamurana.csv — first 20 rows shown
</summary>
```

```{r hamurana-csv-preview}
#| echo: false
# Replace with: read_csv("bcs/inflow_hamurana.csv")
hamurana_preview <- inflow_humurana |> 
  slice_head(n = 6) 

hamurana_preview |>
  kbl(digits = 3, caption = "inflow_Hamurana.csv (excerpt)") |>
  kable_styling(bootstrap_options = c("striped","condensed","hover"),
                full_width = FALSE)
```

```{=html}
</details>
```

### `meteo_glm.csv`

::: {.callout-note appearance="minimal"}
Daily meteorological forcing file sourced from ERA5: shortwave radiation, longwave radiation,
cloud cover, air temperature, relative humidity, wind speed (u/v components), and rainfall.
:::

```{=html}
<details>
<summary style="cursor:pointer; font-weight:600; margin-bottom:0.5em;">
  met_rotorua.csv — column reference
</summary>
```

```{r met-col-ref}
#| echo: false
met_preview <- met |> 
  slice_head(n = 6)

met_preview |>
  kbl(caption = "Preview of GLM met data") |>
  kable_styling(bootstrap_options = c("striped","condensed","hover"),
                full_width = FALSE)
```

```{=html}
</details>
```

### `aed_phyto_pars.csv`

::: {.callout-note appearance="minimal"}
Phytoplankton functional group parameters for the three groups configured in AED.
:::

```{=html}
<details>
<summary style="cursor:pointer; font-weight:600; margin-bottom:0.5em;">
  aed_phyto_pars.csv
</summary>
```

```{r phyto-pars}
#| echo: false

phyto_pars |>
  kbl(caption = "Phytoplankton group parameters (aed_phyto_pars.csv)") |>
  kable_styling(bootstrap_options = c("striped","condensed","hover"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE)
```

```{=html}
</details>
```

:::

---

## Inflows

There are fourteen simulated inflows into Lake Rotorua from 1959-2022, work is 
ongoing to update these with more recent data. 

```{r}
#| fig-cap: "Simulated inflows to Lake Rotorua from the fourteen streams."
#| fig-height: 8

plot_aeme_loads(aeme, date_filter = FALSE)

```

```{r}
#| fig-cap: "Estimated annual loads to Lake Rotorua from the fourteen streams."
#| fig-height: 8

plot_annual_loads_stacked(aeme, date_filter = FALSE)

```


## Simulation outputs {#outputs}


### Water temperature {#temp}

```{r plot-temperature}
#| fig-cap: "Simulated surface and bottom water temperature (2020–2022). Dots = observed (example)."
#| fig-height: 4

plot_output_base(aeme, var_sim = "HYD_temp")
```

Currently the model is underestimating water temperature overall, but particularly in the lower levels.

```{r plot-temperature-residuals}
#| fig-cap: "Model residuals for water temperature."
#| fig-height: 8

plot_resid(aeme, var_sim = "HYD_temp")
```

### Thermal stability {#stability}

```{r plot-schmidt}
#| fig-cap: "Schmidt stability index (J/m²) — indicator of stratification intensity."
#| fig-height: 3.5

plot_output(aeme, var_sim = "HYD_schstb")

```

### Lake level {#level}

```{r plot-level}
#| fig-cap: "Simulated lake water level relative to the Ohau Channel weir crest (279.6 m a.s.l.)."
#| fig-height: 3.5

plot_output(aeme, var_sim = "LKE_lvlwtr", )

```

### Nutrients & chlorophyll-a {#nutrients}

```{r plot-nutrients}
#| fig-cap: "Simulated total phosphorus (TP), total nitrogen (TN), and chlorophyll-a."
#| fig-height: 8

plot_output_base(aeme, var_sim = c("NIT_tn","PHS_tp","PHY_tchla"))

```

### Phytoplankton groups {#phyto-groups}

```{r plot-phytos}
#| fig-cap: "Simulated phytoplankton group biomass (mg/m³) for the three groups configured in AED."
#| fig-height: 4

plot_phytos(aeme, remove_spin_up = FALSE)
```


### Model performance summary {#performance}

```{r performance-stats}
#| echo: false

perf |>
  kbl(digits = c(0,0,2,2,2,0),
      col.names = c("Variable","Period","RMSE","Bias","R²","n"),
      caption = "Summary model performance statistics") |>
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = FALSE) |>
  column_spec(4, color = ifelse(perf$Bias > 0, "#1565C0", "#B71C1C")) |>
  footnote(general = "RMSE and Bias in variable units. Observations from BoPRC monitoring programme.",
           general_title = "Note: ")
```

---

## Session info {#session}

```{=html}
<details close>
<summary style="cursor:pointer; font-weight:600; margin-bottom:0.5em;">
  R session information
</summary>
```

```{r session-info}
#| code-fold: true
#| code-summary: "R session information"
sessionInfo()
```

```{=html}
</details>
```
